---
title: "Week6 Project2 Tidy Data"
author: "Chad Bailey"
date: "October 6, 2018"
---

##Overview
This project continues to practice the tasks of data transformation and data wrangling.  

This is done by taking three "untidy" datasets and using R to "tidy" the data. Specifically this project will focus on using the R packages of tidyr and dplyr.


#### Load necessary packages and define supporting functions
```{r message=FALSE}

library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)

repeat.labels <- function(vec)
{
  tmp <-""
  for(i in 1:length(vec))
  {
    if (vec[i] != "") {tmp <- vec[i]}
    if (vec[i] == "" & i != 1) {vec[i] <- tmp}
    if (vec[i] == "" & i == 1) {vec[i] <- ""}
    next
  }
  vec
}
```


#### Dataset1: Officer Deaths

Load the dataset into R  
  
This dataset is split across two tables.
```{r}
fileLocation_1 <- 'https://raw.githubusercontent.com/ChadRyanBailey/607-Week6-Project2/master/04%20dataset1%20-%20Accidental%20Officer%20Deaths%20-%20modified.csv'

dataset1a <- read.table(file = fileLocation_1, header = TRUE, sep = ',', stringsAsFactors =  FALSE)



#get the second table for this dataset
fileLocation_2 <- 'https://raw.githubusercontent.com/ChadRyanBailey/607-Week6-Project2/master/04%20dataset1%20-%20Felony%20Officer%20Deaths%20-%20modified.csv'

dataset1b <- read.table(file = fileLocation_2, header = TRUE, sep = ',', stringsAsFactors =  FALSE)

```


Check the names/structure
```{r}
names(dataset1a)
names(dataset1b)
```

For each table, add a column to allow records to be identified after they are unioned (i.e., appended/stacked)
```{r}
dataset1a <- dataset1a %>% mutate(Death_Type = "Felony")
dataset1b <- dataset1b %>% mutate(Death_Type = "Accidental")
```

For each table, seperate the values for region, sub-region, state/territory into three distinct columns
```{r}

dataset1a <- dataset1a %>% 
  mutate(Region = ifelse(Area %in% c("NORTHEAST"
                                       , "MIDWEST"
                                       , "SOUTH"
                                       , "WEST"
                                       , "PUERTO RICO AND OTHER OUTLYING AREAS"), Area, "")) %>%
  mutate(Sub_Region = ifelse(Area %in% c("New England"
                                         , "Middle Atlantic"
                                         , "East North Central"
                                         , "West North Central"
                                         , "South Atlantic"
                                         , "East South Central"
                                         , "West South Central"
                                         , "Mountain"
                                         , "Pacific"), Area, "")) %>%
  mutate(Area = ifelse(Region != "", "Region Total", 
                       ifelse(Sub_Region != "", "Sub_Region Total", Area))) %>%
  mutate(Sub_Region = repeat.labels(Sub_Region)) %>%
  mutate(Sub_Region = ifelse(Region != "", "Region Total", Sub_Region)) %>%
  mutate(Region = repeat.labels(Region)) %>%
  mutate(Region = ifelse(Region == "", "Grand Total", Region)) %>%
  mutate(Sub_Region = ifelse(Sub_Region == "", "Grand Total", Sub_Region)) %>%
  mutate(Area = ifelse(Area == "Number of victim officers", "Grand Total", Area)) %>%
  rename(State_Territory = "Area")
  
dataset1b <- dataset1b %>% 
  mutate(Region = ifelse(Area %in% c("NORTHEAST"
                                       , "MIDWEST"
                                       , "SOUTH"
                                       , "WEST"
                                       , "PUERTO RICO AND OTHER OUTLYING AREAS"), Area, "")) %>%
  mutate(Sub_Region = ifelse(Area %in% c("New England"
                                         , "Middle Atlantic"
                                         , "East North Central"
                                         , "West North Central"
                                         , "South Atlantic"
                                         , "East South Central"
                                         , "West South Central"
                                         , "Mountain"
                                         , "Pacific"), Area, "")) %>%
  mutate(Area = ifelse(Region != "", "Region Total", 
                       ifelse(Sub_Region != "", "Sub_Region Total", Area))) %>%
  mutate(Sub_Region = repeat.labels(Sub_Region)) %>%
  mutate(Sub_Region = ifelse(Region != "", "Region Total", Sub_Region)) %>%
  mutate(Region = repeat.labels(Region)) %>%
  mutate(Region = ifelse(Region == "", "Grand Total", Region)) %>%
  mutate(Sub_Region = ifelse(Sub_Region == "", "Grand Total", Sub_Region)) %>%
  mutate(Area = ifelse(Area == "Number of victim officers", "Grand Total", Area)) %>%
  rename(State_Territory = "Area")

```



Union (append/stack) the tables and then check counts for reasonability
```{r}
dataset1 <- union(dataset1a, dataset1b)

dataset1 %>%
  group_by(Death_Type) %>%
  summarise(n=n())

```

As requested, all subtotals are removed
```{r}

dataset1 <- dataset1 %>%
  #drop total rows
  filter(Region != "Grand Total" 
         & !Sub_Region %in% c("Grand Total", "Region Total")
         & !State_Territory %in% c("Grand Total", "Sub_Region Total")) %>%
  #drop total column and re-order columns
  select(Death_Type
         ,Region
         , Sub_Region
         , State_Territory 
         , X2008
         , X2009
         , X2010
         , X2011
         , X2012
         , X2013
         , X2014
         , X2015
         , X2016
         , X2017
         ) %>%
  arrange(Death_Type, Region, Sub_Region, State_Territory)
```

Normalize (gather) the data so that years move from being seperate columns to being rows
```{r}
names(dataset1) = str_replace_all(names(dataset1), "X", "")
dataset1_tidy <- gather(dataset1, "Year", "N_Deaths", 5:14)
```

Denormalize (spread) the data so that each value of Death_Type becomes its own column.  
Also add a total column (across death types)
```{r}
dataset1_tidy <- spread(dataset1_tidy, Death_Type, N_Deaths)

dataset1_tidy <- dataset1_tidy %>%
  mutate(Total = Felony + Accidental)

```

Filter to a single state to verify transformations were succuessful
```{r}
dataset1_tidy %>%
  filter(State_Territory == "Michigan") %>%
  arrange(Region, Sub_Region, State_Territory)

```

Calculate some basic statistics
```{r}

dataset1_summary <- dataset1_tidy %>%
  group_by(Year, Region) %>%
  summarise(nRecords = n()
            ,nTotalDeaths = sum(Total)
            ,nAccidentalDeaths = sum(Accidental)
            ,nFelonyDeaths = sum(Felony))

```


Generate a scatterplot for visual exploration
```{r}
g <- ggplot(dataset1_summary, aes(x= nAccidentalDeaths, y = nFelonyDeaths)) 
g <- g + geom_point(aes(color = Region))
g <- g + facet_wrap(~Year)
g


```


#### Dataset2: US Race and Ethnicity Counts

Load the dataset into R
```{r}
fileLocation <- 'https://raw.githubusercontent.com/ChadRyanBailey/607-Week6-Project2/master/06%20dataset2%20-%20Race%20Counts%20-%20modified.csv'

dataset2 <- read.table(file = fileLocation, header = TRUE, sep = ',', stringsAsFactors =  FALSE)
```

Check the names and structure
```{r}
names(dataset2)
head(dataset2)
```

Begin tidying up the dataset.

1. Add a new column for "Count of Races" to hold current subtotals
2. Add change the "Race" value for these subtotal rows
3. Make sure row labels are repeated for each row
4. Replace the character "2" with the word "Two" for consistency
5. Rename columns for easier reference 

```{r}
dataset2 <- dataset2 %>%
  mutate(Count_of_Races = ifelse(Race %in% c("All races", "One race", "2+ races"), Race, "")) %>%
  mutate(Race = ifelse(Count_of_Races != "", "All races", Race)) %>%
  mutate(Count_of_Races = repeat.labels(Count_of_Races)) %>%
  mutate(Count_of_Races = str_replace_all(Count_of_Races, "2", "Two")) %>%
  mutate(Race = str_replace_all(Race, "2", "Two")) %>%
  rename(H_L = "Hispanic.or.Latino"
         ,Pct_of_H_L = "X..of.H.L"
         ,Pct_H_L_of_US = "X..of.US"
         ,Not_H_L = "Not.Hispanic.or.Latino"
         ,Pct_of_Not_H_L = "X..of.Not.H.L"
         ,Pct_Not_H_L_of_US = "X..of.US.1"
         )

```

Add columns for an "All Ethnicities" block of columns

1. Add column for race to be totaled across ethnicities
2. Add column for race pct of ethnicity
3. Get US grand total
4. Calculate the Pct of US total for each race total
```{r}
dataset2 <- dataset2 %>%
  mutate(RaceTotal = H_L + Not_H_L) %>%
  mutate(RacePct_ofEth = 100)

#get US grand total to calculate RaceTotal Pct of US
us_gt <-as.numeric(dataset2 %>%
  filter(Count_of_Races == "All races") %>%
  select(RaceTotal))

dataset2 <- dataset2 %>%
  mutate(RacePct_ofUS = round(RaceTotal/us_gt*100, 2))
```

Seperate the three block of columns to seperate dataframes.  
Then union (append/stack) those dataframes together.

```{r}
#get just Hispanic records
eth_Hispanic <- dataset2 %>%
  select(Count_of_Races ,Race ,H_L ,Pct_of_H_L ,Pct_H_L_of_US) %>%
  mutate (Ethnicity = "Hispanic or Latino") %>%
  rename(N_Race_Ethnicity = "H_L"
         , Pct_Ethnicity_Total = "Pct_of_H_L"
         , Pct_US_Total = "Pct_H_L_of_US") %>%
  mutate(Pct_US_Total = as.character(Pct_US_Total))

#get just not-hispanic records
eth_Not_Hispanic <- dataset2 %>%
  select(Count_of_Races ,Race ,Not_H_L ,Pct_of_Not_H_L ,Pct_Not_H_L_of_US) %>%
  mutate (Ethnicity = "Not Hispanic or Latino") %>%
  rename(N_Race_Ethnicity = "Not_H_L"
         , Pct_Ethnicity_Total = "Pct_of_Not_H_L"
         , Pct_US_Total = "Pct_Not_H_L_of_US") %>%
  mutate(Pct_US_Total = as.character(Pct_US_Total))

#get just "all ethnicities" records
eth_Total <- dataset2 %>%
  select(Count_of_Races ,Race ,RaceTotal ,RacePct_ofEth ,RacePct_ofUS) %>%
  mutate (Ethnicity = "All Ethnicities") %>%
  rename(N_Race_Ethnicity = "RaceTotal"
         , Pct_Ethnicity_Total = "RacePct_ofEth"
         , Pct_US_Total = "RacePct_ofUS") %>%
  mutate(Pct_US_Total = as.character(Pct_US_Total))

# union (i.e., append/stack) the three datasets together
dataset2_tidy <- union(eth_Total, eth_Hispanic)
dataset2_tidy <- union(dataset2_tidy, eth_Not_Hispanic)

```
Check the count of records for each block (Ethnicity)
```{r}
dataset2_tidy %>%
  group_by(Ethnicity) %>%
  summarise(n = n())
```

In some records the Pct_US_Total is shown as "<0.1" rather than as a true percent. The following uses the available data to recalculate the Pct_US_Total and rounds to two decimals to avoid this problem.  
  
The same is done for Pct_Ethnicity_Total so the two variables will have the same precision.
```{r}

#US grand total is already stored in us_gt

#get grand totals for each Ethnicity
eth_totals <- dataset2_tidy %>%
  filter(Count_of_Races == "All races") %>%
  rename(eth_Total = "N_Race_Ethnicity") %>%
  select(Ethnicity, eth_Total)

#join ethnicity grand totals into dataset for calculations
dataset2_tidy <- left_join(dataset2_tidy, eth_totals, by = "Ethnicity")

#recalculate and stored pct columns
dataset2_tidy <- dataset2_tidy %>%
  mutate(Pct_Ethnicity_Total = round(N_Race_Ethnicity/eth_Total*100, 2)) %>%
  mutate(Pct_US_Total = round(N_Race_Ethnicity/us_gt*100, 2)) %>%
  select(Ethnicity
         , Count_of_Races
         , Race
         , N_Race_Ethnicity
         , Pct_Ethnicity_Total
         , Pct_US_Total) %>%
  arrange(Ethnicity
         , Count_of_Races
         , Race)

dataset2_tidy %>% 
  filter(Ethnicity == "All Ethnicities")

```



#### Dataset3: US Poverty by Various Demographics

Load the data into R
```{r}

fileLocation <- 'https://raw.githubusercontent.com/ChadRyanBailey/607-Week6-Project2/master/07%20dataset3%20-%20Poverty%20Counts.csv'

dataset3 <- read.table(file = fileLocation, header = TRUE, sep = ',')
```

Review the column names
```{r}
names(dataset3)
```

The column names are particularly long and cumbersome. Adjust column names to be more workable.
```{r}

names(dataset3) <- str_replace_all(names(dataset3), "\\.", "_")
names(dataset3) <- str_replace_all(names(dataset3), "Poverty", "Pov")
names(dataset3) <- str_replace_all(names(dataset3), "Number", "N")
names(dataset3) <- str_replace_all(names(dataset3), "Percent", "Pct")
names(dataset3) <- str_replace_all(names(dataset3), "_Below", "")
names(dataset3) <- str_replace_all(names(dataset3), "_Margin_of_Error", "_MOE")
names(dataset3) <- str_replace_all(names(dataset3), "Total", "Total_N")



names(dataset3)
```

Add a Characterist_Type variable to group the types of characteristics  

```{r}

dataset3 <- dataset3 %>%
  mutate(Characteristic = str_trim(Characteristic)) %>%
  mutate(Characteristic_Type = ifelse(is.na(X2016_Total_N), Characteristic, "")) %>%
  mutate(Characteristic_Type = repeat.labels(Characteristic_Type)) %>%
  mutate(Characteristic_Type = str_replace_all(Characteristic_Type, " and Hispanic Origin", "/Ethnicity")) %>%
  mutate(Characteristic_Type = str_replace_all(Characteristic_Type, "\\d", "")) %>%
  filter(X2016_Total_N != "" #remove blank rows
         & Characteristic != "Total"  #remove total row
         & Characteristic != "White, not Hispanic") #remove sub-element row

head(dataset3 %>% select(Characteristic_Type, Characteristic), n = 15)
```

Seperate year-block of columns to its own dataframe and then union together.
```{r}
#get just 2016 block of records
y_2016 <- dataset3 %>%
  mutate(X2016_Year = 2016) %>%
  select( X2016_Year
         , Characteristic_Type
         , Characteristic
         , X2016_Total_N
         , X2016_Pov_N
         , X2016_Pov_N_MOE
         , X2016_Pov_Pct
         , X2016_Pov_Pct_MOE)

names(y_2016) <- str_replace_all(names(y_2016), "X2016_", "")


#get just 2017 block of records
y_2017 <- dataset3 %>%
  mutate(X2017_Year = 2017) %>%
  select( X2017_Year
         , Characteristic_Type
         , Characteristic
         , X2017_Total_N
         , X2017_Pov_N
         , X2017_Pov_N_MOE
         , X2017_Pov_Pct
         , X2017_Pov_Pct_MOE)

names(y_2017) <- str_replace_all(names(y_2017), "X2017_", "")
  
#union (append/stack) the years back together
dataset3_tidy <- union(y_2016, y_2017)
```

The orginal table was showing counts in thousands. The following recalculates to the actual values.
```{r}
dataset3_tidy <- dataset3_tidy %>%
  mutate(Total_N = Total_N *1000, Pov_N = Pov_N*1000)
```

Review a small sample to verify the transformations were successful.
```{r}
dataset3_tidy %>%
  filter(Characteristic_Type %in% c("Race/Ethnicity", "Sex")) %>%
  select(Year ,Characteristic_Type ,Characteristic ,Total_N ,Pov_N, Pov_Pct) %>%
  arrange (Year ,Characteristic_Type ,Characteristic) 

```
