---
title: "607 Week9 Assignment"
author: "Chad Bailey"
date: "October 28, 2018"
---

##Overview
This assignment explores the process of using web APIs to aquire data and then load that data into an R dataframe.

## Get the data
```{r message=FALSE}

# load necessary supporting packages
library(httr)
library(plyr)
library(dplyr)

# get the parts of the api request
url_base <-'http://api.nytimes.com/svc/semantic/v2/geocodes/'
query_header <- '/query.json?'
query_condition1 <- 'country_code=US'
api_key_header <- "api_key="
api_key <- readLines('C:/Users/cbailey/Desktop/CUNY MSDS/607 Data Aquisition and Management/Week9/nytimes_api_key.csv')

# construct api request and stored the results
response <- GET(paste(url_base
                      ,query_header
                      ,query_condition1
                      ,'&'
                      ,api_key_header
                      ,api_key
                      ,sep = ''))
 
# parse the api results
response_parsed <- content(response, "parse")

# element 4 of the parsed results contains the data of interest
# unlist element 4, which produces a list of lists
unlisted <- sapply(response_parsed[[4]], unlist)

# transpose each of the second-layer lists
unlisted_trans <- lapply(unlisted, t)

# load each of the transposed second-layer lists as dataframes
dfs <- lapply(unlisted_trans , data.frame , stringsAsFactors = FALSE)

# bind the individual dataframes (records) together as a single dataframe 
# filling in any missing values with NA
df_response <- rbind.fill(dfs)

#check names of the database columns
names(df_response)

#tidy the data
df_response_slim <- df_response %>% 
  mutate(longitude = as.numeric(longitude)
         ,latitude = as.numeric(latitude)
         ,population = as.numeric(population)) %>%
  rename(state_code = admin_code1) %>%
  select (state_code, name, latitude, longitude)

df_response_slim

```



##Plot the Geographic Points
```{r message=FALSE, warning=FALSE}

#load libraries needed for map plotting
library(ggplot2)
library(ggmap)

# load the USA states outlines
states <- map_data("state")
ggplot(data = states) + 
  geom_polygon(aes(x = long, y = lat, group = group)
                 , fill = "lightblue"
                 , color = "black") + 
  
  coord_fixed(1.3) +      #adjusting scales for proper display
    guides(fill=FALSE) +  #turn off color legends
  theme_nothing() +       #turn off grid and axes
  geom_point(data = df_response_slim
             , aes(x = longitude, y = latitude)
             , color = "red"
             , size = 2)  #add api geo points

```

##Conclusions
This project demonstrated one method of acquiring web API data and a rather simple way to load it into an R dataframe. 

Lastly, since the data acquired was geographic data, the information was plotted on a map of the USA.



