---
title: "607 Week5 Assignment"
author: "Chad Bailey"
date: "September 28, 2018"
---


##Overview
This assignment explores the tasks of data transformation and aggregations within the R environment.  

This is done using the provided "Flights Arrival Delay" dataset and the tidyr and dplyr packages.


## Load and review the data
Load the .csv file
```{r}

fileLocation <- 'https://raw.githubusercontent.com/ChadRyanBailey/607-Week5-Assignment/master/02%20Airlines%20Arrival%20and%20Delays.csv'

flightsData <- read.table(file = fileLocation, header = TRUE, sep = ",")

head(flightsData)
```

## Form possible questions
This will help guide how the data will need to be restrutured.  

The dataset appears to be about two airlines' flight status counts for five cities. A few possible question would be: 

1. Which airline is larger (i.e., has more total flights)?
2. Which airline has a better overall rate of "on time" flights?
3. Does either airline have a better rate of "on time" flights for all cities?  

## Perform necessary cleanup and transformations

Adjust column names
```{r}
names(flightsData) <- c("Airline"
                        ,"Flight_Status"
                        ,"Los_Angeles"
                        ,"Phoenix"
                        ,"San_Diego"
                        ,"San_Francisco"
                        ,"Seattle")
```

Subset to rows with data and reset row numbers
```{r}
keepRows <-flightsData$Flight_Status != ''

flightsData2 <- flightsData[keepRows,]
rownames(flightsData2) <- NULL
```

Transform the Airline column so that each row lists its airline name
```{r}
airline <- flightsData2$Airline
for(i in 1:length(airline))
{
  if (airline[i] != '') {tmp <- airline[i]}
  if (airline[i] == '') {airline[i] <- tmp}
  next
}
flightsData2$Airline <- airline
```

Get check the headers and first few rows of the altered table
```{r}
head(flightsData2)

```

Transforrm the table from wide to long to with two new columsn [City] and [n_Flights]
```{r}

library(tidyr)
flights_long <- gather(flightsData2, "City", "n_Flights", 3:7)
head(flights_long)

```

Transform the table so that each value of Flight_Status becomes its own column. This will allow for easier comparison of flight statuses. 
```{r}

flights_long$Flight_Status <- gsub(" ", "_", flights_long$Flight_Status)

flights_tidy <- spread(flights_long, Flight_Status, n_Flights)

head(flights_tidy)
```

Add new columns to help with analysis.
```{r message=FALSE}

library(dplyr)
flights_tidy <- mutate(flights_tidy
                       , total = delayed + on_time
                       , pct_on_time = on_time / total * 100)

head(flights_tidy)

```

## Analyze and compare the data
Calculate the overall on-time rate for each airline. Do this both as grand total rollup (weighted average) and a straight (unweighted) average of the variable pct_on_time.
```{r}
airlines_overall <- flights_tidy %>% 
  group_by(Airline) %>% 
  summarise(total = sum(total)
            , pct_on_time_rollup = sum(on_time) / sum(total) * 100 
            , pct_on_time_avg = mean(pct_on_time))

airlines_overall

```

The different methods of aggregation do not provide aligned results.  

Visually explore the transformed datasets. Given the large imbalance in the two airlines' total flights, be sure to consider this variable in the exploration. 

```{r}
library(ggplot2)

overall <- ggplot(airlines_overall, aes(x=pct_on_time_avg, y = pct_on_time_rollup))
overall + geom_point(aes(color = Airline, size = total))
```

Explore if the imbalance of the number of flights is consistent across cities
```{r}
by_city <- ggplot(flights_tidy, aes(x=total, y = pct_on_time)) 
by_city + geom_point(aes(color = Airline, size = total)) + facet_wrap(~City)
```

It appears Alaska has a higher on_time rate in all five cities included in the dataset. However, AM WEST has significantly more flights and has its largest number of flights at Phoenix, where its on_time rate is highest. This hides AM WEST's lower on_time rates in other cites.  

Show this finding through additional table transformations.
```{r}
pct_on_time <- flights_tidy %>% 
                select (Airline, City, pct_on_time) %>%
                spread(Airline , pct_on_time)

total <- flights_tidy %>% 
           select (Airline, City, total) %>%
           spread(Airline , total)

airline_rates <- inner_join(total, pct_on_time, by = "City")

names(airline_rates) <- c("City"
                          ,"Alaska_n"
                          ,"AM_WEST_n"
                          ,"Alaska_pct_ontime"
                          ,"AM_WEST_pct_ontime")

airline_rates %>%
  select (City, Alaska_n, Alaska_pct_ontime, AM_WEST_n, AM_WEST_pct_ontime)

```


##Conclusion
While AM WEST is by far the larger airline, having nearly double the total number of flights, passengers are more likely to have an on time flight with Alaska regardless of which city is being considered.
