---
title: "607 Data Acquistion & Management Week 2 Assignment"
author: "Chad Bailey"
date:   "`r Sys.Date()`"
---



##Overview

This assignment explores moving data across multiple environments, specifically .csv files, MySQL, and R.

The assignment processes through the following steps:

0. Obtain new Data
1. Load new data into a MySQL database
2. Join/transform the data in MySQL
3. Export the final dataset from MySQL to a .csv file
4. Load the .csv file to R


All files used for this assignment are stored at https://github.com/ChadRyanBailey/607-Week2-Assignment

##0. Obtain new data
In this particular case:

1. Choose six recent popular movies
2. For each movie produce at least 5 records, each having a 1-5 rating
3. Save this data in a .csv file "03 Movie Ratings.csv"

##1. Load new data into a MySQL database
Use the MySQL script, "04 Load Movie Ratings.sql", to load the .csv file into the MySQL database

##2. Join/Transform the data in MySQL
No transformations are needed for this assignment

##3. Export the final dataset from MySQL to a .csv file
Use the MySQL script, "05 Export Movie Ratings.sql", to export the data to a new .csv file "06 Movie Ratings SQL Output.csv"

##4. Load the .csv file to R
Load the Data:
```{r}
fileLocation <- "C:/ProgramData/MySQL/MySQL Server 8.0/Uploads/Data 607/Week2/06 Movie Ratings SQL Output.csv"
MovieRatings <- read.table(file = fileLocation, header = TRUE, sep = ",")
```

Review the loaded data:
```{r}
head(MovieRatings)
summary(MovieRatings)
```

 



 


