---
title: "607 Week4 Project1"
author: "Chad Bailey"
date: "September 21, 2018"
output: html_document
---



##Project Summary
This project explores the ability to perform basic data wrangling functions with R: 

  * Load data into R and exporting data from R
  * Using regular expressions to extract portions of larger text strings
  * Transforming data orientation
  * Changing data types
  * Joining/merging dataframes
  * Performing aggregations on dataframes

The desired task is to use R to take a provided semi-structured text file containing information on chess ratings and output a fully-structured .csv file with the following information for all records:

* Player's Id
* Player's Name
* Player's State
* Player's Total Number of Points
* Player's Pre-Rating
* Average Opponent's Pre-Rating

In this context a semi-structured file is one that is structured to the human eye but is not laid out in the machine readable structure of rows and columns. Whereas a fully-structured file is a laid out in a machine readable fashion.


## Load the text file and the necessary R packages
```{r}

library(stringr)
library(plyr)
library(reshape2)

input_file_name <- "https://raw.githubusercontent.com/ChadRyanBailey/607-Week4-Project1/master/02%20tournamentinfo.txt"

textFile <- readLines(input_file_name, warn = FALSE)
```

## Explore the text file
```{r}
head(textFile, n=10)
tail(textFile, n=10)
```

It appears the text file is laid out with the first three rows containing file header information and the last row being a visual divider not containing data.

The remaining rows are in three block groupings:

  1. A row having a visual divider with no data
  2. A row with data visually partitioned
  3. A another row, of the same record, visually partitioned

## Transform the data into a machine-readable layout
Start by removing the header and last rows as they do not follow the same pattern as other rows. Then within each 3-block of rows load each row with data (rows 2 and 3) to its own vector.

```{r}
#drop the headers and the final row 
textFileNRows <- length(textFile)
rows_with_pattern <- textFile [c(4:(textFileNRows-1))]

#In 3-block rows: ignore row1 (visual divider) and load rows 2 and 3 to seperate vectors
record_row1 <- rows_with_pattern[seq(from = 2, to = length(rows_with_pattern), by = 3)]
record_row2 <- rows_with_pattern[seq(from = 3, to = length(rows_with_pattern), by = 3)]
```

Then beginning with the row1 vector, within each row split apart the character string to individual elements.
```{r}
#For row1 split apart to individual elements and then load into a dataframe
data_elements <- unlist(str_split(record_row1, "\\|"))
data_elements <- str_trim(data_elements) 
de_matrix <- matrix(data_elements, nrow = length(record_row1), byrow = TRUE)
de_matrix <- de_matrix[, 1:(ncol(de_matrix)-1)]
de_dataframe <- data.frame(de_matrix, stringsAsFactors = FALSE)    
names(de_dataframe) <- c("Id"
                         , "Name"
                         , "Total_Points"
                         , "Round1"
                         , "Round2"
                         , "Round3"
                         , "Round4"
                         , "Round5"
                         , "Round6"
                         , "Round7")

#Convert column data types as needed
de_dataframe["Id"] <- apply(de_dataframe["Id"], MARGIN = c(1), FUN = as.numeric)
de_dataframe["Total_Points"] <- apply(de_dataframe["Total_Points"], MARGIN = c(1), FUN = as.numeric)

head(de_dataframe)
```

The data for each round contain two pieces of information (status and Opponent Id) that still need to be split apart.  

First get the outcome status of the round.

```{r}
#limit to columns with rounds data
rounds <- de_dataframe[, 4:10]

#get just the rounds status
round_status_func <- function(data) 
{
  unlist(str_extract_all(data, "[[:alpha:]]"))
}
round_status <- data.frame(apply(rounds, MARGIN = c(1,2), FUN = round_status_func), stringsAsFactors =  FALSE)

names(round_status) <- c("R1_Status"
                         , "R2_Status"
                         , "R3_Status"
                         , "R4_Status"
                         , "R5_Status"
                         , "R6_Status"
                         , "R7_Status")
```

Then get the opponent Id for the round.
```{r}
#get just the round's OppoentId
round_opp_func <- function(data) 
{
  test <- unlist(str_extract_all(data, "[[:digit:]]{1,}"))
  if (length(test) == 0) {NA}
  else {test}
}

round_oppId <- apply(rounds, MARGIN = c(1,2), FUN = round_opp_func)
round_oppId <- apply(round_oppId, MARGIN = c(1,2), FUN = as.numeric)
round_oppId <- data.frame(round_oppId, stringsAsFactors = FALSE)

names(round_oppId) <- c("R1_Opp_Id"
                         , "R2_Opp_Id"
                         , "R3_Opp_Id"
                         , "R4_Opp_Id"
                         , "R5_Opp_Id"
                         , "R6_Opp_Id"
                         , "R7_Opp_Id")
```


Moving on the the row2 vector, within each row split apart the character string to individual elements. Because most of the data in the second row is not needed for this project a more direct approach for extracting the individual elements is taken.
```{r}

head(record_row2)

State <- str_trim(unlist(str_extract_all(record_row2, "\\s\\w{2}\\s")))

Pre_Rating <- unlist(str_extract_all(record_row2, "R:[ ]{0,}\\d{1,}"))
Pre_Rating <- unlist(str_extract_all(Pre_Rating, "\\d{1,}"))
Pre_Rating <- as.numeric(Pre_Rating)

```

Pull the transformed elements back together.
```{r}

tournament <- cbind(de_dataframe[, c("Id", "Name", "Total_Points")]
                    , round_status
                    , round_oppId
                    , State
                    , Pre_Rating)

head(tournament)
```

##Determine the average pre-ratings of opponents
This requires another data transformation: melting (i.e., unpivoting) the opponent Ids so that for each player Id gets a block of 7 rows - one for each round's opponent Id.

```{r}
opponentsDenorm <- tournament[, c("Id"
                            , "R1_Opp_Id"
                            , "R2_Opp_Id"
                            , "R3_Opp_Id"
                            , "R4_Opp_Id"
                            , "R5_Opp_Id"
                            , "R6_Opp_Id"
                            , "R7_Opp_Id")]

OpponentsNorm <- melt(opponentsDenorm, id.vars = c("Id"), variable.name = "Round", value.name = "OpponentId")

OpponentsNorm$Round <- unlist(str_extract_all(OpponentsNorm$Round, "\\d"))
OpponentsNorm$Round <- as.numeric(OpponentsNorm$Round)

head(OpponentsNorm[OpponentsNorm$Id == 1, ])
```

Then merge/join the transformed tournament dataframe to the transformed opponent dataframe
```{r}
oppPreRatings <- merge(x=OpponentsNorm, y=tournament[,c("Id", "Pre_Rating")]
      ,by.x=c("OpponentId")
      ,by.y=c("Id"))

head(oppPreRatings[oppPreRatings$Id == 1, ], n=7)
```

Finally, calculate the average (mean) opponent pre-rating 
```{r}
avgOppPreRating <- aggregate(Pre_Rating ~ Id, oppPreRatings, mean, na.action = na.omit)
names(avgOppPreRating) <- c("Id", "Avg_Opp_Pre_Rating")
avgOppPreRating$Avg_Opp_Pre_Rating <- round(avgOppPreRating$Avg_Opp_Pre_Rating, 0)

head(avgOppPreRating)
```

## Construct the desired output dataset
First merge/join the average opponent pre-rating with the transformed tournament dataframe.
```{r}

tournament <- oppPreRatings <- merge(x=tournament, y=avgOppPreRating
      ,by.x=c("Id")
      ,by.y=c("Id"))
```

Then limit to the desired columns/fields.
```{r}
tournamentSubset <- tournament[, c("Id"
                                   , "Name"
                                   , "State"
                                   , "Total_Points"
                                   , "Pre_Rating"
                                   , "Avg_Opp_Pre_Rating")]

head(tournamentSubset)

```

##Output the data to a .csv file

```{r}

output_file_name <- "C:/Users/cbailey/Desktop/tournamentSubset.csv"

write.csv(tournamentSubset, file = output_file_name, row.names = FALSE)
```






